Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic

### 2022-10-17 manuscript.Rmd
second draft with edits

author: 
  - name          : "Samantha L. Heiman"
    affiliation   : "1"
    equal_contrib : yes
  - name          : "Scott Claessens"
    affiliation   : "2"
    equal_contrib : yes
  - name          : "Edward R. Hurt"
    affiliation   : "1"
  - name          : "Peter M. Todd"
    affiliation   : "1,3"
    corresponding : yes
    address       : "1101 E 10th St, Bloomington, IN 47405, United States"
    email         : "pmtodd@indiana.edu"

affiliation:
  - id            : "1"
    institution   : "Department of Psychological and Brain Sciences, Indiana University Bloomington, United States"
  - id            : "2"
    institution   : "School of Psychology, University of Auckland, New Zealand"
  - id            : "3"
    institution   : "Cognitive Science Program, Indiana University Bloomington, United States"

abstract: |
 Social norms allow humans to coordinate and cooperate in the face of existential threats. In particular, injunctive social norms prescribe what people _ought_ to do, whereas descriptive social norms inform what people _actually_ do. While previous experimental work has revealed people’s sensitivity to normative influence, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their influences on cooperative behavior over time. To understand how social norms emerge and shape behavior in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked people’s reported mask wearing behavior and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior are tightly coupled and both change quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that descriptive norms, but not injunctive norms, caused future increases in mask wearing behavior. These findings underline the relative importance of descriptive social norms in shaping cooperative behavior during uncertain times.

note              : "This working paper has not yet been peer-reviewed. This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation."
keywords          : "descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "xxxx words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

@@

Despite decades of research on the causes and consequences of injunctive and descriptive norms [@Bicchieri2009; @Cialdini1991; @Cialdini2007; @Cialdini2021; @Schultz2007], social norms have remained an elusive concept in the behavioral sciences and several open questions remain [@Fehr2018a; @VanKleef2019]. In the current work, we focus on two such questions. First, how do injunctive and descriptive norms emerge naturally over time within a population? Second, how do evolving injunctive and descriptive norms influence behavior?

@@

With regards to normative influences on behavior, there is a wealth of cross-sectional evidence demonstrating the behavioral impact of social norms. For example, field experiments have demonstrated the positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. Evidence also suggests that any potentially deleterious effects of descriptive social norms (e.g., choosing to litter at a picnic site that already contains visible signs of littering) can be counteracted by instead focusing individuals’ attention on injunctive norms [@Schultz2007]. However, these cross-sectional studies have two main limitations. First, studies have not adequately controlled for other potential non-social influences on behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014]. For example, descriptive social norms might only affect cooperative behavior indirectly by influencing perceptions of the effectiveness of the behavior (i.e., factual beliefs) or individuals' moral preferences (i.e., personal normative beliefs). Second, cross-sectional studies have tended to follow experimental designs in which perceptions of social norms are manipulated by the researchers at a single time point, and thus do not allow cooperative social norms to emerge and affect behavior naturally over time within a population. An alternative way to identify causality, whilst retaining ecological validity, is to follow perceptions of social norms and cooperative behavior over time amidst a real, unfolding social dilemma. This is particularly important because social norms are not static: they change dynamically over time through processes of deliberation and social interaction [@Smith2015].

To understand how novel injunctive and descriptive social norms emerge over time and shape cooperative behavior in a non-experimental setting, we studied mask wearing behavior during the COVID-19 pandemic. Before the pandemic, mask wearing was not a common behavior in the United States. In April 2020, two months into the pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) as a cooperative behavior that people should adopt to minimize the spread of COVID-19. But mask wearing posed a social dilemma to individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the wider community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the evolution of mask wearing in the United States throughout the COVID-19 pandemic allows us to study, on a short timescale within a single population, the emergence of novel injunctive and descriptive social norms and the causal effects of these norms on cooperative behavior.

Recent research has found positive relationships between perceptions of social norms and protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between between-person and within-person change over time [@Hamaker2015], nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the entire pandemic. These studies also do not control for the influences of non-social beliefs, such as factual beliefs and personal normative beliefs [@Bicchieri2014].

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. We aimed to answer two main research questions. First, how do descriptive and injunctive mask wearing norms emerge and evolve over time in the United States population? Second, how do descriptive and injunctive mask wearing norms affect mask wearing behavior over time?

@@

To understand how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration. Figure \@ref(fig:plotTimeline) plots self-reported mask wearing behavior and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum’s COVID-19 Timeline [@CDC2022b] and are included here based on their direct relevance to protective behaviors (e.g., changing CDC guidance on mask wearing, vaccination access, etc.).

(ref:plotTimelineCaption) _Timeline of self-reported mask wearing behavior and perceived social norms in the United States throughout the COVID-19 pandemic._ (a) Points and line ranges indicate mean averages and standard errors for the self-reported mask wearing item. This item was measured across all eighteen time points on a 5-point Likert scale, with higher values indicating increased frequency of personal mask wearing during in-person interactions. (b) Points and line ranges indicate mean averages and standard errors for perceived descriptive mask wearing norms (blue) and perceived injunctive mask wearing norms (green). These items were measured across eleven time points on a 7-point Likert scale, with higher values indicating stronger perceived social norms. (c) Smoothed data for new COVID-19 cases in the United States, displayed on the log scale (data retrieved from Our World in Data; https://ourworldindata.org/). Across all panels, gray dashed lines represent significant pandemic-related events in the United States, such as vaccine approval from the Food and Drug Administration (FDA) and public health recommendations from the Centers for Disease Control and Prevention (CDC).

@@

Two main observations can be made about the emergence and stability of social norms from these visualizations. First, social norms and behavior were tightly coupled over time. Although social norms are measured on fewer occasions than mask wearing behavior, we can see that as mask wearing behavior decreased in the summer of 2021, so too did perceived descriptive and injunctive mask wearing norms. Subsequently, the steep rise in COVID-19 case numbers in the fall of 2021 saw concomitant increases in both mask wearing behavior and perceived social norms, though to lower levels than before. In line with these patterns, multilevel regression models revealed positive correlations between mask wearing behavior and perceived descriptive mask wearing norms (_b_ = `r printnum(fixef(m1.1)["DesNorms"])`, 95% confidence interval [`r printnum(conf1.1["DesNorms","2.5 %"])` `r printnum(conf1.1["DesNorms","97.5 %"])`]) and between mask wearing behavior and perceived injunctive mask wearing norms (_b_ = `r printnum(fixef(m1.2)["InjNorms"])`, 95% CI [`r printnum(conf1.2["InjNorms","2.5 %"])` `r printnum(conf1.2["InjNorms","97.5 %"])`]) across individuals and time points (Supplementary Figure \@ref(fig:plotCorBehNorm); Supplementary Table \@ref(tab:modelSummaryTable1)).

Second, fluctuations in mask wearing behavior and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing behavior subsequently declined after the CDC rescinded their mask wearing recommendation in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased restrictions in March 2022. Impressively, shifts in mask wearing behavior and norms happened on the timescale of weeks. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \@ref(fig:plotCDCSens); Supplementary Table \@ref(tab:changePointsTable)).

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing behavior and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing behavior, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from trait levels for our main variables and controls: perceived descriptive mask wearing norms, perceived injunctive mask wearing norms, factual beliefs, personal normative beliefs, and self-reported mask wearing behavior (see Supplementary Figure \@ref(fig:plotDAG) for justification of control variables). According to established fit statistics, this model fitted the data well (root mean square error of approximation = `r printnum(fitMeasures['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures['rmsea.ci.upper'], digits = 3)`]; standardized root mean squared residual = `r printnum(fitMeasures['srmr'], digits = 3)`; comparative fit index = `r printnum(fitMeasures['cfi'], digits = 3)`). Since we are primarily interested in the causal effects of social norms on behavior, we focus on these variables in what follows (but see Supplementary Table \@ref(tab:lavaanTable) for full list of estimated autoregressive and cross-lagged effects).

@@

(ref:plotRICLPMCaption) _Results of ten-wave unconstrained random-intercept cross-lagged panel model._ Arrows represent within-person autoregressive and cross-lagged effects from the model, partitioning out stable between-person individual differences and controlling for factual beliefs, personal normative beliefs, and political orientation. Arrow thickness is scaled according to standardized effect size. Bolded arrows indicate significantly positive parameters, _p_ < 0.05. Gray arrows indicate non-significant parameters.

@@

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing behavior. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing behavior. According to recent effect size guidelines for cross-lagged panel models [@Orth2022], the standardized beta coefficients for these cross-lagged effects were large (Time 2, $\beta$ = `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.lower"])` `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_05" & pe$op == "~" & pe$rhs == "w3_02","pvalue"])`; Time 5, $\beta$ = `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.lower"])` `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_09" & pe$op == "~" & pe$rhs == "w3_05","pvalue"])`; Time 11, $\beta$ = `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.lower"])` `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_13" & pe$op == "~" & pe$rhs == "w3_11","pvalue"])`; Time 13, $\beta$ = `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.lower"])` `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_14" & pe$op == "~" & pe$rhs == "w3_13","pvalue"])`). These cross-lagged effects from descriptive norms to mask wearing diminish in 2022. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing behavior predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

@@

In our longitudinal analysis, we found that _descriptive_ norms, not injunctive norms, independently predicted future increases in mask wearing. In line with this finding, descriptive norms have also been shown to predict future increases in physical distancing and prosocial behaviors (e.g., neighborhood help, charitable donations) throughout the COVID-19 pandemic, though this previous work did not adequately disentangle between-person and within-person effects [@Rudert2021]. Similarly, experimental work has revealed that people in the United States are more likely to report intentions to wear a mask if they are told that others are wearing masks [@Bokemper2021]. Why this specific effect of descriptive norms on mask wearing? Descriptive norms are particularly useful for fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic [@Gelfand2015]. During times of uncertainty, people look to others to quickly coordinate their behavior, and attempt to alleviate uncertainty-related stress by identifying with their group and its social norms [@Hogg2008; @Wellen1998]. Supporting this situational-uncertainty explanation, our model revealed that descriptive norms predicted future mask wearing at the beginning of the pandemic, when uncertainty was likely at its height. Unsure of how to respond to a novel threatening situation, people began to look to their neighbors and adapted their mask wearing behavior accordingly. However, as the uncertainty surrounding the pandemic lessened in 2022, reliance on descriptive norms appeared to decrease.

>>>>

author: 
  - name          : "Samantha L. Heiman"
    affiliation   : "1"
    equal_contrib : yes
    corresponding : yes
    address       : "1101 E 10th St, Bloomington, IN 47405, United States"
    email         : "slheiman@iu.edu"
  - name          : "Scott Claessens"
    affiliation   : "2"
    equal_contrib : yes
  - name          : "Jessica D. Ayers"
    affiliation   : "3"
  - name          : "Diego Guevara Beltran"
    affiliation   : "3"
  - name          : "Andrew Van Horn"
    affiliation   : "4"
  - name          : "Edward R. Hirt"
    affiliation   : "1"
  - name          : "Athena Aktipis"
    affiliation   : "3"
  - name          : "Peter M. Todd"
    affiliation   : "1,5"

affiliation:
  - id            : "1"
    institution   : "Department of Psychological and Brain Sciences, Indiana University Bloomington, United States"
  - id            : "2"
    institution   : "School of Psychology, University of Auckland, New Zealand"
  - id            : "3"
    institution   : "Department of Psychology, Arizona State University, United States"
  - id            : "4"
    institution   : "Department of Anthropology, Rutgers University - New Brunswick, United States"
  - id            : "5"
    institution   : "Cognitive Science Program, Indiana University Bloomington, United States"

abstract: |
 Social norms allow humans to coordinate and cooperate in the face of existential threats. In particular, injunctive social norms prescribe what people _ought_ to do, whereas descriptive social norms inform what people _actually_ do. While previous experimental work has revealed people’s sensitivity to normative influence, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their causal influences on cooperative behavior over time. To understand how social norms emerge and shape behavior in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked people’s reported mask wearing behavior and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior were tightly coupled and both changed quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that descriptive norms, but not injunctive norms, caused future increases in mask wearing behavior. These findings underline the relative importance of descriptive social norms in shaping cooperative behavior during uncertain times.

note              : "This working paper has not yet been peer-reviewed. This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation."
keywords          : "descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "4683 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

@@

Despite decades of research on the causes and consequences of injunctive and descriptive norms [@Bicchieri2009; @Cialdini1991; @Cialdini2007; @Cialdini2021; @Schultz2007], social norms have remained an elusive concept in the behavioral sciences and several open questions remain [@Fehr2018a; @Legros2020; @VanKleef2019]. In the current work, we focus on two such questions. First, how do injunctive and descriptive norms emerge naturally over time within a population? Second, how do evolving injunctive and descriptive norms causally influence behavior over time?

@@

With regards to normative influences on behavior, there is a wealth of cross-sectional evidence demonstrating the behavioral impact of social norms. For example, field experiments have demonstrated the positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. Evidence also suggests that any potentially deleterious effects of descriptive social norms (e.g., choosing to litter at a picnic site that already contains visible signs of littering) can be counteracted by instead focusing individuals’ attention on injunctive norms [@Schultz2007]. However, these cross-sectional studies have two main limitations. First, studies have not adequately controlled for other potential non-social influences on behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014]. For example, social norms might only affect cooperative behavior indirectly by influencing perceptions of the effectiveness of the behavior (i.e., factual beliefs) or individuals' moral preferences (i.e., personal normative beliefs). Second, cross-sectional studies have tended to follow experimental designs in which perceptions of social norms are manipulated by the researchers at a single time point, and thus do not allow cooperative social norms to emerge and affect behavior naturally over time within a population. An alternative way to identify causality, whilst retaining ecological validity, is to follow perceptions of social norms and cooperative behavior over time amidst a real, unfolding social dilemma. This is particularly important because social norms are not static: they change dynamically over time through processes of deliberation and social interaction [@Smith2015].

To understand how novel injunctive and descriptive social norms emerge over time and shape cooperative behavior in a non-experimental setting, we studied mask wearing behavior during the COVID-19 pandemic. Before the pandemic, mask wearing was not a common behavior in the United States. In April 2020, two months into the pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) as a cooperative behavior that people should adopt to minimize the spread of COVID-19 [@CDC2022]. Mask wearing has individual benefits, but the CDC emphasized the cooperative aspects of mask wearing as a tool for reducing community spread [@CDC2020]. Indeed, mask wearing posed a social dilemma to individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the wider community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the evolution of mask wearing in the United States throughout the COVID-19 pandemic allows us to study, on a short timescale within a single population, the emergence of novel injunctive and descriptive social norms and the causal effects of these norms on cooperative behavior.

Recent research has found positive relationships between perceptions of social norms and protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between within-person and between-person change over time [@Hamaker2015], nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the entire pandemic. These studies also do not adequately distinguish between the effects of social norms and the effects of other non-social beliefs that are also predicted to shape behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014].

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Between September 2020 and October 2022, we asked participants to report their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. We used these data to answer two main research questions. First, how do descriptive and injunctive mask wearing norms emerge and evolve over time in the United States population? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing behavior over time?

@@

To understand how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration. Figure \@ref(fig:plotTimeline) plots self-reported mask wearing behavior and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum’s COVID-19 Timeline [@CDC2022] and are included here based on their direct relevance to protective behaviors (e.g., changing CDC guidance on mask wearing, vaccination access, etc.).

(ref:plotTimelineCaption) _Timeline of self-reported mask wearing behavior and perceived social norms in the United States throughout the COVID-19 pandemic._ (a) Points and line ranges indicate mean averages and standard errors for the self-reported mask wearing item. This item was measured across all eighteen time points on a 5-point Likert scale, with higher values indicating increased frequency of personal mask wearing during in-person interactions. (b) Points and line ranges indicate mean averages and standard errors for perceived injunctive mask wearing norms (green) and perceived descriptive mask wearing norms (blue). These items were measured across eleven time points on a 7-point Likert scale, with higher values indicating stronger perceived social norms. (c) Smoothed data for new COVID-19 cases in the United States, displayed on the log scale (data retrieved from Our World in Data; https://ourworldindata.org/). Across all panels, gray dashed lines represent significant pandemic-related events in the United States, such as vaccine approval from the Food and Drug Administration (FDA) and public health recommendations from the Centers for Disease Control and Prevention (CDC).

@@

Two main observations can be made about the emergence and stability of social norms from these visualizations. First, social norms and behavior were tightly coupled over time. Although social norms are measured on fewer occasions than mask wearing behavior, we can see that as mask wearing behavior decreased in the summer of 2021, so too did perceived descriptive and injunctive mask wearing norms. Subsequently, the steep rise in COVID-19 case numbers in the fall of 2021 saw concomitant increases in both mask wearing behavior and perceived social norms, before declining again in 2022. In line with these patterns, multilevel regression models revealed positive correlations between mask wearing behavior and perceived descriptive mask wearing norms (_b_ = `r printnum(fixef(m1.1)["DesNorms"])`, 95% confidence interval [`r printnum(conf1.1["DesNorms","2.5 %"])` `r printnum(conf1.1["DesNorms","97.5 %"])`], _p_ `r printp(summary(m1.1)$coefficients["DesNorms","Pr(>|t|)"])`) and between mask wearing behavior and perceived injunctive mask wearing norms (_b_ = `r printnum(fixef(m1.2)["InjNorms"])`, 95% CI [`r printnum(conf1.2["InjNorms","2.5 %"])` `r printnum(conf1.2["InjNorms","97.5 %"])`], _p_ `r printp(summary(m1.2)$coefficients["InjNorms","Pr(>|t|)"])`) across individuals and time points (Supplementary Figure \@ref(fig:plotCorBehNorm); Supplementary Table \@ref(tab:modelSummaryTable1)).

Second, fluctuations in mask wearing behavior and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing behavior subsequently declined after the CDC rescinded their mask wearing recommendation following widespread vaccine availability in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased mask wearing guidelines in March 2022. Impressively, shifts in mask wearing behavior and norms happened on the timescale of weeks. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \@ref(fig:plotCDCSens); Supplementary Table \@ref(tab:changePointsTable)).

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing behavior and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing behavior, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from trait levels for our main variables (self-reported mask wearing behavior, perceived descriptive mask wearing norms, and perceived injunctive mask wearing norms) and time-variant control variables (factual beliefs and personal normative beliefs). In line with our proposed causal model (Supplementary Figure \@ref(fig:plotDAG)), we also included political orientation as an exogenous time-invariant control. According to established fit statistics, this model fitted the data well (root mean square error of approximation = `r printnum(fitMeasures['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures['rmsea.ci.upper'], digits = 3)`]; standardized root mean squared residual = `r printnum(fitMeasures['srmr'], digits = 3)`; comparative fit index = `r printnum(fitMeasures['cfi'], digits = 3)`). Since we are primarily interested in the causal effects of social norms on behavior, in what follows we focus on the results for mask wearing behavior, perceived descriptive norms, and perceived injunctive norms (but see Supplementary Table \@ref(tab:lavaanTable) for full list of estimated autoregressive and cross-lagged effects).

@@

(ref:plotRICLPMCaption) _Results of ten-wave unconstrained random-intercept cross-lagged panel model._ Arrows represent within-person autoregressive and cross-lagged effects for mask wearing behavior and perceived descriptive and injunctive norms, partitioning out stable between-person individual differences and controlling for factual beliefs, personal normative beliefs, and political orientation. Arrow thickness is scaled according to standardized effect size. Bolded arrows indicate significantly positive parameters, _p_ < 0.05. Gray arrows indicate non-significant parameters.

@@

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing behavior. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing behavior. According to recent effect size guidelines for cross-lagged panel models [@Orth2022], the standardized beta coefficients for these cross-lagged effects were large (Time 2, $\beta$ = `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.lower"])` `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_05" & pe$op == "~" & pe$rhs == "w3_02","pvalue"])`; Time 5, $\beta$ = `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.lower"])` `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_09" & pe$op == "~" & pe$rhs == "w3_05","pvalue"])`; Time 11, $\beta$ = `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.lower"])` `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_13" & pe$op == "~" & pe$rhs == "w3_11","pvalue"])`; Time 13, $\beta$ = `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.lower"])` `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_14" & pe$op == "~" & pe$rhs == "w3_13","pvalue"])`). In 2022, these cross-lagged effects from descriptive norms to mask wearing behavior diminished. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing behavior predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

@@

In our longitudinal analysis, we found that descriptive norms, not injunctive norms, predicted future increases in mask wearing across multiple time points. These cross-lagged effects were independent of the effects of non-social beliefs and political orientation. In line with this finding, descriptive norms have also been shown to predict future increases in physical distancing and prosocial behaviors (e.g., neighborhood help, charitable donations) throughout the COVID-19 pandemic, though this previous work did not adequately disentangle between-person and within-person effects [@Rudert2021]. Similarly, experimental work has revealed that people in the United States are more likely to report intentions to wear a mask if they are told that others are wearing masks [@Bokemper2021]. Why this specific effect of descriptive norms on mask wearing? Descriptive norms are particularly useful for fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic [@Gelfand2015]. During times of uncertainty, people look to others to quickly coordinate their behavior, and attempt to alleviate uncertainty-related stress by identifying with their group and its social norms [@Hogg2008; @Wellen1998]. Supporting this situational-uncertainty explanation, our model revealed that descriptive norms predicted future mask wearing at the beginning of the pandemic, when uncertainty was likely at its peak. Unsure of how to respond to a novel threatening situation, people began to look to their neighbors and adapted their mask wearing behavior accordingly. However, as the uncertainty surrounding the pandemic lessened in 2022, reliance on descriptive norms appeared to decrease.
